thesis2024

### 2025-03-24 thesis2024.tex
First render

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\chapter{Within and Beyond: Chert Procurement Patterns During The Upper
Palaeolithic in Southwesternmost
Iberia}\label{within-and-beyond-chert-procurement-patterns-during-the-upper-palaeolithic-in-southwesternmost-iberia}

>>>>

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\chapter{Within and beyond: chert procurement patterns during the Upper
Palaeolithic in Southwesternmost
Iberia}\label{within-and-beyond-chert-procurement-patterns-during-the-upper-palaeolithic-in-southwesternmost-iberia}

\chapter{Within and beyond: chert procurement patterns during the Upper
Palaeolithic in Southwesternmost
Iberia}\label{within-and-beyond-chert-procurement-patterns-during-the-upper-palaeolithic-in-southwesternmost-iberia-1}

\textbf{Authors}: Joana Belmiro, Xavier Terradas, Salvador
Dominguez-Bella, João Cascalheira

\newpage

\section*{Abstract}\label{abstract-1}
\addcontentsline{toc}{section}{Abstract}

\markright{Abstract}

Analyses of raw materials and the distinction between local/regional and
long-distance sources have proven invaluable for understanding the
extensive movements, interactions, and social networks during the Upper
Palaeolithic in the Iberian Peninsula. However, unlike other parts of
Iberia, research on the management and acquisition of raw materials in
the south and west of Iberia remains relatively underdeveloped. Despite
significant knowledge about the technological practices of Palaeolithic
hunter-gatherers from southern Portugal, particularly from studies
conducted at the site of Vale Boi, there is a noticeable lack of focus
on raw materials management. This paper presents the first comprehensive
characterization of chert raw materials from the Gravettian,
Proto-Solutrean, and Solutrean occupations at Vale Boi, using both
macroscopic and petrographic techniques. Our study reveals that the
majority of chert found at Vale Boi originates locally, within a 20 km
radius. However, a non-negligible portion of the chert comes from
non-local sources, indicating \textgreater200 km raw material
circulation from central Portugal and southern Spain.

\textbf{keywords:} Upper Paleolithic; Iberian Peninsula; Lithic raw
materials; Petrography

\newpage

\section{Introduction}\label{introduction-1}

Knappable raw materials play a crucial role in understanding the
mobility and lifeways of past hunter-gatherers, given the ubiquitous
presence of lithic artefacts throughout Prehistory. More than mere rocks
with which stone tools were produced, the management of these resources
is intimately connected to a group's technological, social, and cultural
organisation, potentially influencing the group's overall survivability
(Oestmo 2017; Binford 1979; Bleed 1986; Bousman 1993; Torrence 1983;
Gould and Saggers 1985).

Several key topics in the study of hunter-gatherer behaviour and
organisation have been explored in different geographic regions and
chronological periods through raw material analysis, such as modalities
of procurement and mobility strategies (Ambrose and Lorenz 1990; Binford
1979; Binford and Stone 1985; Gould 1985; Gould and Saggers 1985; Kuhn
1991; McCall 2007), occupation types and their duration (Kuhn 2004;
Surovell 2009), the establishment and dimension of social networks
(Whallon 2006), as well as exchanges between groups or individuals
(Gamble 1999). These analyses are often achieved by systematically
characterizing geological and archaeological raw materials and
establishing correlations between samples from both origins. The
ultimate aim is to identify the corresponding sources, enabling, for
example, a precise differentiation between local and non-local raw
materials, and examine their distribution within a site over different
occupations and periods.

These approaches and concepts have also been extensively and
successfully applied to the study of lithic assemblages from prehistoric
archaeological sites in the Iberian Peninsula (e.g., Aubry et al. 2016;
Aubry and Igreja 2009; García-Rojas et al. 2021; Gómez de Soler et al.
2020; Herrero-Alonso, Fuertes-Prieto, and Neira-Campos 2020; Matias
2016; Ortega 2003; Pereira, Andrade, et al. 2016; Pereira et al. 2021,
2022; Ramacciotti et al. 2022; Sánchez de la Torre et al. 2023; Soto
2016; Rodríguez, Rodríguez, and Pelegrin 2011; Costa et al. 2022; Nocete
et al. 2005), and contributed to understanding how different groups
explored and managed the available lithic resources.

In this context, raw material analyses and the distinction between
local, regional and long-distance sources have been instrumental in
tracing extensive movements, contacts, and social networks throughout
the Upper Palaeolithic (UP) throughout Iberia. Noteworthy examples
include studies in northwestern Iberia (Galicia) during the Solutrean
period (Hermida, Rellán, and Rodríguez 2016), northern Iberia (Asturias)
during the Magdalenian occupations at Las Caldas (Corchón Rodríguez,
Martínez, and Tarriño 2016), northeastern Iberia (Catalunya) during the
Aurignacian and Gravettian periods at Arbreda cave (Marreiros et al.
2016; Ortega 2003), northern Portugal during the Gravettian/Solutrean
periods in the Côa Valley (Aubry et al. 2016, 2004), and inland central
Iberia (Guadalajara) during the Proto-Solutrean and Solutrean periods at
Peña Capón (Sánchez de la Torre et al. 2023). Ultimately, the
recognition of these connections has broadened our comprehension of
mobility and social organization during the UP in Iberia, illustrating a
vast interconnected territory.

However, in contrast to other areas of Iberia, and despite the potential
of raw material studies to shed light on the lifeways and organisational
structures of Palaeolithic hunter-gatherers, research on the procurement
and management of raw materials during the UP in southern and western
Iberia remains in its initial stages. Specifically, in southern
Portugal, even though there is considerable knowledge about the
technological organisation of Palaeolithic hunter-gatherers---largely
due to research conducted at the Vale Boi archaeological site (see e.g.,
Belmiro et al. 2021; Nuno Bicho et al. 2013; Cascalheira and Bicho 2013;
Marreiros et al. 2015; Cascalheira 2010; Horta, Cascalheira, and Bicho
2019)---raw material studies are notably infrequent and use mainly
macroscopic methodologies.

This situation is attributed to the absence of a detailed and
comprehensive characterization of regional lithic resources in the
Algarve region (southern Portugal) and the subsequent lack of a complete
reference collection produced by complementing characterization
methodologies, which is a critical component for conducting raw material
studies. Such efforts are particularly crucial for distinguishing
between local and non-local raw materials and for understanding
variations within the archaeological record (Pop 2015; Surovell 2009).

Recently, we have made significant strides by introducing such a
reference collection and establishing a framework for raw material
studies in the region, through a detailed macroscopic and petrographic
database of chert resources in the Algarve region (Belmiro, Terradas,
and Cascalheira 2023). This initiative paves the way for systematic
investigations into raw material procurement strategies during the Upper
Palaeolithic, especially concerning the Vale Boi site, where previous
raw material studies have already shown complex patterns of chert
procurement and management. These preliminary studies at Vale Boi,
employing only macroscopic methods, pointed to a predominant use of
local cherts (Nuno Bicho et al. 2013; Pereira, Bicho, et al. 2016;
Verissimo 2004), likely originating within a 16 km radius and reflecting
embedded procurement strategies (Pereira, Bicho, et al. 2016). To a
lesser extent, non-local cherts were also identified throughout the
various UP occupations (Nuno Bicho et al. 2013; Pereira, Bicho, et al.
2016; Verissimo 2004), pointing to extended sourcing networks, possibly
involving sourcing from other regions of southern Portugal and southern
Spain. The presence of chert from the Cretaceous formations of central
Portugal has also been suggested (Nuno Bicho et al. 2013).

These results highlight the potential for systematic raw material
studies at the site to shed light on the lithic resource procurement
strategies, mobility patterns, and social networks of the
hunter-gatherer groups of southwestern Iberia throughout the UP.

Vale Boi stands as a key archaeological site for this study since it is
currently the only site in southern Portugal with a long-term occupation
spanning most of the UP and allowing the exploration of raw material
procurement and mobility trends through time (Cascalheira et al. 2017).
It also holds a pivotal role in the study of UP adaptations on the
Iberian Peninsula, providing invaluable insights into the subsistence
and technological strategies of hunter-gatherer populations (Belmiro et
al. 2021; Nuno Bicho et al. 2013; Cascalheira et al. 2017; Cascalheira
2019; Horta, Cascalheira, and Bicho 2019; Manne et al. 2012; Marreiros
et al. 2016; Pereira, Bicho, et al. 2016).

Furthermore, previous studies have highlighted the role of Vale Boi in
understanding different patterns of territory exploitation and contact
with other regions of Iberia throughout the UP. During the Solutrean,
Vale Boi has been interpreted as an important connection point between
the core Solutrean areas of central Portugal and southern Spain, making
it a key site for identifying extended social networks that promoted the
exchange of information and culture (Cascalheira et al. 2017;
Cascalheira and Bicho 2013; Cascalheira 2013). On the contrary,
techno-typological studies from the Gravettian occupations of Vale Boi
instead show marked differences between regions and suggest a more
limited circulation of people and information in southern Iberia
(Marreiros and Bicho 2013).

Based on the notions that a) during the Solutrean Vale Boi served as a
connection point between different regions which may have promoted the
exchange of information and culture, including raw materials from
long-distances; and b) the limited circulation of people during the
Gravettian may have isolated groups occupying Vale Boi, making them
mostly reliant on the available local or regional resources, we propose
a set of hypotheses and expectations regarding chert procurement and
human mobility throughout the UP at Vale Boi. On one hand, it would be
expected that Solutrean occupations are characterized by the predominant
use of local raw materials but with a considerable percentage of cherts
obtained from long-distance sources, possibly through exchange with
different groups in Iberia. Conversely, it would be expected that
Gravettian occupations would be mostly composed of local cherts, with a
more limited presence of long-distance cherts.

To investigate the applicability of these patterns to the procurement
and circulation of lithic raw materials, and to examine the prevailing
hypotheses concerning UP mobility and social network exchanges, this
paper focuses on the following questions: 1) what types of chert were
the hunter-gatherers of Vale Boi using and where did they come from?; 2)
are there different patterns of chert use throughout time?; 3) and if
so, how can these be related to patterns of land use and the circulation
of people/ideas, as suggested through technological studies of the same
lithic assemblages? As a result, our study presents the first
comprehensive characterisation of chert raw materials, through
macroscopic and microscopic analysis, of UP occupations in southwestern
Iberia, derived from the Gravettian, Proto-Solutrean, and Solutrean
(c.~32-19 ka cal BP) occupations at Vale Boi.

\subsection{Site description}\label{site-description}

Vale Boi is located on the western coast of the Algarve region in
southern Portugal, within a small valley that extends southward to the
Atlantic coast, approximately 2 km away (Figure~\ref{fig-map} b). It is
bordered by limestone outcrops that form rock shelters facing west and
southwest (Nuno Bicho et al. 2007; Nuno Bicho, Cascalheira, and
Marreiros 2012; Nuno Bicho et al. 2013; Cascalheira and Bicho 2013;
Cascalheira et al. 2017; Manne et al. 2012; Manne and Bicho 2011). The
site extends for more than 10,000 sq meters along the slope of the
valley, through which three main areas were excavated between 2000 and
2019: Slope, Terrace, and Shelter. For this study, the chosen areas were
the Terrace and Shelter areas (see Section 2 Materials and Methods).

The Terrace encompasses occupations from the UP to the Early Neolithic,
resulting in the identification of eight main lithostratigraphic units.
The UP sequence includes several occupations attributed to the
Gravettian (levels 8 to 6) between c.~32 and 27 ka cal BP,
Proto-Solutrean (levels 5 to 4E) between c.~26 and 24 ka cal BP, and
Solutrean (level 4D, 4C, 4C, 4 and Lower 3) between c.~24 and 20 ka cal
BP (Belmiro et al. 2021; Cascalheira and Bicho 2013; Cascalheira et al.
2017).

The Shelter area shows four main lithostratigraphic units, with
Magdalenian, Solutrean and Gravettian occupations. From these, the
Solutrean levels show the most intensive occupation, with three,
well-preserved, archaeological horizons (layers C to A) and dated to
c.~24-22 ka cal BP (Cascalheira et al. 2013, 2017; Cascalheira 2010).
These occupation levels were identified under blocks of limestone, which
collapsed from the rockshelter ceiling (Cascalheira 2010).

Both these areas have been previously interpreted as seasonal
residential camps, repeatedly used for extended stays, due to the
abundance of lithic debitage, stone tools, heat-cracked rocks related to
grease rendering activities, large quantities of faunal remains (both
marine and terrestrial) and the presence of ornaments and portable art
(Manne et al. 2012). The analysis of lithic assemblages and retouched
frequency of lithic assemblages from the Terrace and Shelter areas also
corroborate this interpretation (Cascalheira 2010; Marreiros 2009;
Cascalheira et al. 2017). Cascalheira et al. (2017) show that, with the
exception of the Early Gravettian occupations of the Terrace area and
Magdalenian occupation of the Shelter, all other UP occupations are
composed of high-density assemblages with low degree of retouch,
correspondent to a residential base-camp occupation, for extended
periods of time.

\includegraphics{figures/within-and-beyond/map_layout.png}

\caption{\label{fig-map}Iberian Peninsula map with locations and
geological formations mentioned throughout the paper. a) Detail of the
central Portugal region where the Cenomanian formations with chert
nodules are located. b) Detail of southwestern Portugal, with chert and
chalcedony bearing formations, outcrops, and location of Vale Boi. c)
Detail of eastern Algarve with chert-bearing formations. The figure was
produced using ArcGIS Pro 3.2.}

\subsection{Regional geological
background}\label{regional-geological-background}

The Algarve region is a complex territory, marked by several geomorphic
sub-regions and geological units that provide different lithological
resources. Two of the main geological units which characterize the
Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the south
Portuguese Zone (Fernandes et al. 2012; Oliveira 1984; Onézime et al.
2003), located in the north sector of the region, and characterized by
schists, greywackes and quartzites; and 2) the Algarve basin in the
south, which is characterized by limestones, dolomites and marls
(Terrinha et al. 2006). It is within these limestones and dolomitized
limestones from the Triassic to the Upper Jurassic where chalcedony and
chert nodules/beds can be found (Manuppella et al. 1987; Manuppella et
al. 2007; Ribeiro 2005; Rocha et al. 1979; Rocha et al. 1983; Terrinha
et al. 2006).

The availability of regional resources containing cherts and the
characterization of their lithologies has already been published by our
team (Belmiro, Terradas, and Cascalheira 2023), and we herein present
the summarized main findings. The results showed the existence of
abundant chert nodules and beds found in the Lower Jurassic (Carixian)
formation and Upper Jurassic (Kimmeridgian) formation, mostly
outcropping in the cliff and beach areas of westernmost Algarve, at
\textasciitilde20 km southwest from the site of Vale Boi, and with one
single inland outcrop located \textasciitilde8 km east from the site
(Figure~\ref{fig-map} b). The Lower Jurassic nodules and beds can be
found in abundant outcrops, but also frequently in sub-primary or
secondary deposition settings in proximity to the outcrops, possibly due
to natural erosion. However, in the case of the inland outcrop of
Ferrel, the outcrop is partially destroyed due to recent human action,
and chert can be found scattered as chunks or larger nodules. The Upper
Jurassic nodules can be found in sub-primary deposition and are easily
accessible, or in secondary deposition. The cherts from the previously
mentioned formations are are frequently opaque with yellow, grey, purple
or red colours, occasionally with bands or laminations, and frequently
with visible sponge spicules. Although abundant, these chert nodules are
frequently small (\textasciitilde4-10 cm in length), with the largest
nodules being \textasciitilde20 cm, and frequently characterized by high
levels of dolomitization and fractures. Petrographically, they are
composed mainly of microcrystalline quartz, with wackestone or packstone
textures, small percentages of macrocrystalline quartz, fibrous
chalcedony and allochems composed of oxides and poorly preserved
bioclasts. When identifiable, the bioclasts are Echinoderms,
Radiolarians, Sponge spicules and Bivalve shells. The Upper Jurassic
cherts of Mareta show similar petrographic characteristics, but with
abundant Calcispheres which are not always apparent under the
stereomicroscope.

Chert sources were also identified in two formations from the eastern
area of the Algarve (Figure~\ref{fig-map} c). The Middle Jurassic Malhão
formation chert outcrops were identified at around 80-100 km east of the
site and characterized by chert nodules with similar macroscopic
features to those found in western Algarve. Petrographically they are
composed mainly of microcrystalline quartz and characterized by
wackestone textures, while dolomite, chalcedony and macrocrystalline
quartz are also present in small percentages (\textless10\%). Allochems
are iron oxides, and albeit poorly preserved, several bioclasts can be
identified: Sponge spicules, Radiolarians, Ostracods, Echinoderms,
Calcispheres and possibly Tentaculites. The Upper Jurassic Jordana
formation chert outcrops were identified at around 90 km east of Vale
Boi and characterized by chert nodules with significantly different
macroscopic features to those of other formations, with frequently small
sizes (\textasciitilde5 cm), difficult removal from the parent rock and
unidentified secondary deposits. Petrographically they are composed
mostly of microcrystalline quartz, with wackestone to packstone
textures, and small percentages of fibrous chalcedony and dolomite
(\textless10\%). They show frequent iron oxides and very frequent
bioclasts, albeit often poorly preserved and unidentifiable; identified
fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods,
Echinoderms and Gastropods.

Chalcedony can be found \textasciitilde10 km west of Vale Boi
(Figure~\ref{fig-map} b), within the marl-carbonated Triassic formations
as lenses or nodules and was formed through hydrothermal processes (Nuno
Bicho et al. 2013). Previous studies also note the presence of
chalcedony (with frequent presence of cortex) in the immediate vicinity
of the site (Pereira, Bicho, et al. 2016; Gibaja Bao and Ferreira Bicho
2013). The chalcedony frequently has fractures and fissures and is
characterized by a fibrous quartz structure, without inclusions or
fossils.

\section{Materials and methods}\label{materials-and-methods}

To characterize the siliceous raw materials used during the UP
occupations at Vale Boi and to identify their geological sources, we
focused on chalcedony and chert archaeological materials from the
Terrace and Shelter areas. These two areas were chosen since both have
detailed spatial information, well-defined chronological sequences, and
previous studies providing lithic technology, faunal, and
geoarchaeological data. For this, we focused on layers that spanned the
UP occupations, characterized by good preservation, a significant number
of lithic finds and with radiocarbon dating (as mentioned in section
Site description).

In this study, the terms chert and chalcedony follow the definitions
provided by Luedtke (1992). We use the term chert to refer to
sedimentary rocks composed primarily of microcrystalline quartz, which
includes other possible varieties such as jasper, agate, or chalcedony.
The term chalcedony is used in its petrographic sense and refers to
cherts with a fibrous quartz structure. Since this definition of
chalcedony implies a destructive method which cannot be applied to all
archaeological chalcedony samples, we use the term based on the
confirmed correlation between the macroscopic appearance of chalcedony
and its petrographic characteristics, obtained through geological and
archaeological thin sections.

Chert was chosen for this study as it is one of the three main raw
materials used at the archaeological site of Vale Boi alongside quartz
and greywacke. Other raw materials such as chalcedony, dolerite, and
schists have also been identified (Belmiro et al. 2021; Cascalheira
2010; Pereira, Bicho, et al. 2016).

Using the field database with all individually coordinated lithic
artefacts (all pieces larger than \textasciitilde2 cm; (Belmiro 2020)),
we compared the incidence of chert and chalcedony in opposition to the
remaining lithics of other raw materials. Figure~\ref{fig-rm-comparison}
shows that for the selected levels of the Terrace area
(Figure~\ref{fig-rm-comparison} a), chert represents only
\textasciitilde15-20\% of the lithic assemblage, while in the Shelter
area (Figure~\ref{fig-rm-comparison} b), chert represents
\textasciitilde35-50\% of the lithic assemblage from the Solutrean
layers A, B, and C. Based on previous lithic studies in the Terrace and
Shelter areas, the remaining 80\% of other raw materials are mostly
composed of quartz and greywacke (Belmiro et al. 2021; Cascalheira 2010;
Marreiros 2009).

Although systematic raw material research has not been conducted on
these materials, according to the regional geological context, they have
been interpreted as local and readily accessible in the vicinity of the
site (Pereira, Bicho, et al. 2016). Frequently, the quartz and greywacke
assemblages include large amounts of shattered and unknapped chunks or
slabs, and have been interpreted in great part as functionally
specialised in activities related to the fragmentation of bones in the
case of quartz and greywack (Cascalheira et al. 2017), or in grease
rendering in the case of quartz (Manne et al. 2012). Previous studies
from the Terrace area have also shown that when removing shatter and
chunks, the percentages of chert become much more representative when
compared to greywacke and quartz (Belmiro et al. 2021; Belmiro 2020).
Based on this, despite the smaller amounts of individually plotted chert
artefacts in the field database, chert continues to be one of the main
raw materials to produce lithic stone tools, with complete knapping
sequences and formal toolkits. The importance of chalcedony can also be
seen in its use to produce formal tool kits during specific
technocomplexes at the site, such as a Vale Comprido point, the
fossil-director for the Proto-Solutrean in western Portugal (Belmiro et
al. 2021) or a laurel leaf preform in the Solutrean layers (Gibaja Bao
and Ferreira Bicho 2013).

\includegraphics{article2_files/figure-pdf/fig-rm-comparison-1.pdf}

\caption{\label{fig-rm-comparison}Comparison of chert, chalcedony, and
other raw materials piece plotted in the Terrace (a) and Shelter (b)
areas of Vale Boi.}

We employed both macroscopic and petrographic methods to achieve a more
comprehensive characterization of the materials, aiming to offset the
inherent limitations of each method (Luedtke 1992).

Macroscopic analysis, while revealing several limitations related to the
absence of quantitative variables (Bustillo et al. 2009), proved to be
advantageous given the extensive sample size of chalcedony and chert
lithics under examination (n=4458). The distinct features of various
local cherts facilitate their differentiation from cherts of
neighbouring origins. Consequently, a macroscopic approach offered a
non-destructive, efficient, and cost-effective method for characterizing
all materials (Bustillo et al. 2009; Tarriño and Terradas 2013).
Therefore, we implemented an initial phase of macroscopic analysis,
subsequently augmented by petrographic study to diminish the
subjectivity of the analysis and enhance the material characterization
with petrographic data. The synergistic application of these two methods
has been previously validated in the characterization of Algarve region
cherts, following recent geological survey efforts (Belmiro, Terradas,
and Cascalheira 2023). Hence, an extensive macroscopic and petrographic
collection was already available for comparative analysis with
archaeological materials, employing consistent methodologies. This
reference collection is the Lusolit lithotheque, hosted physically at
ICArEHB (University of Algarve, Faro) and digitally at
www.lusolit.icarehb.com, and aided in the characterization and source
attribution of cherts.

Hand samples were used throughout the macroscopic characterization for
comparison with the archaeological materials, and the archaeological
thin sections were compared with the previously studied geological chert
thin sections. To identify the source of other lithologies which are not
congruent with the regional cherts and identify possible long-distance
chert procurement, we used hand samples and thin sections of cherts from
central Portugal also available in the Lusolit lithotheque. The
lithotheque from the Unit of Geoarchaeology and Archaeometry Applied to
Historic Artistic and Monumental Heritage from the University of Cadiz
(UCA) was also visited to understand the macroscopic variability of the
local and regional cherts of the lower Guadalquivir basin, within the
Atlantic strip of southwestern Spain (present-day provinces of Huelva
and Cádiz).

Following previous studies, we classify the distance of chert sources
from the site as local (1-30 km from the site), regional (30-120 km),
and long-distance (\textgreater120 km) (Herrero-Alonso, Fuertes-Prieto,
and Neira-Campos 2020; Tarriño, Elorrieta, and García-Rojas 2015).

The macroscopic analysis of the archaeological materials was divided
into two steps. The first step included a preliminary characterization
using a hand lens of 10x magnification. This step allowed us to create
types and subtypes based on similar macroscopic characteristics
frequently acknowledged to be useful to describe and discern between
different types of chert; these included characteristics such as colour,
translucency, and feel (Luedtke 1992)-the complete data dictionary for
the recorded variables can be found in the Supplementary Information
(Online Resource 1). In the second step, we used a Nikon SMZ25
stereomicroscope to observe each sample in further detail, which allowed
us to better characterize the artefacts (especially regarding inclusions
and alterations), and the previously established types - the data
dictionary used for the individual samples can also be found in the
Supplementary Information (Online Resource 1). Given the inherent
macroscopic variability of the geological samples and archaeological
specimens, as well as the number of samples, the database used for the
individual analysis was simplified and focused on collecting data
related to weight, cortex, and alterations. This first step was applied
to a total of 4458 artefacts: a sample of 3627 chert artefacts from the
UP sequence of the Terrace area (levels 7 to 4) and 831 chert artefacts
from the Solutrean sequence of the Shelter area (layers A, B, C; all
samples without an attributed layer on the database were excluded from
the analysis).

While the advantages of employing a macroscopic methodology for
analysing the cherts of Vale Boi are notable, a limitation of this
approach is its inability to unequivocally characterize and categorize
artefacts that exhibit surface alterations typical of an open-air site,
including those related to fire, since weathered and altered surface may
destroy visible structures and even alter the chemical composition
(Delluniversità et al. 2019). Other studies frequently exclude samples
that display intense fire damage or several surface alterations
(Delluniversità et al. 2019; Gómez de Soler et al. 2020; Soto 2016). For
the present study, whenever artefacts displayed extensive alterations
and could not be unequivocally characterized
(Figure~\ref{fig-terrace-alterations}), and given the inexistence of a
reference collection for local altered cherts, they were analysed using
the individual sample dataset (Online Resource 1) but grouped in a
category of indeterminate cherts (INDET). This category included key
alterations such as heat/fire-related alterations which clearly impacted
the macroscopic characteristics of the samples, fully changing its
color, luster and texture, and covering them in crazings (e.g.,
Figure~\ref{fig-terrace-alterations} a). Similarly, it also included
alteration rinds that extended through most of the sample, impeding the
identification of its original colour, texture, translucency, inclusions
and fossils (e.g., Figure~\ref{fig-terrace-alterations} b) and the
presence of extended pitting. However, whenever surface alterations were
present in only a limited area of the sample leaving recognizable
macroscopic features and fossils, or when key elements were
distinguishable to make a characterization (e.g., reddening in the Type
2 cherts, Online Resource 2, figure s5), the samples were not collapsed
into the INDET category.

As seen in Table~\ref{tbl-indet-freq}, level 4 shows the highest
percentages of indeterminate artefacts, corresponding to
\textasciitilde40\% of the total sample (n=1037). This is due to the
frequent alterations of the artifacts, especially related to patinas and
alteration rinds, pits, and fire/heat alterations. For all other levels
in the Terrace, indeterminate represents on average \textasciitilde28\%
of the chert artefacts, while in the Shelter the frequency of INDET is
slightly lower (\textasciitilde20-25\%).

\includegraphics{figures/within-and-beyond/terrace_alterations.png}

\caption{\label{fig-terrace-alterations}Samples with extensive
alterations which prevented the identification of chert type. a)
artefact with cracks, white patina and reddening covering 100\% of the
surface; b) artefact with surface alterations, displaying a portion with
patina (brown) and a portion without patina (grey), cleaned with
Hydrochloric acid.}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0947}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0769}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 22\tabcolsep) * \real{0.0828}}@{}}

\caption{\label{tbl-indet-freq}Number and percentage of grouped and
indeterminate (INDET) artefacts by level in the Terrace (levels 4 to 7)
and Shelter (A to C) areas.}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4C
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4D
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4E
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4E/5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
6
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
7
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
A
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
B
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
C
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Grouped & 619 (59.69\%) & 287 (71.75\%) & 287 (66.74\%) & 332 (73.45\%)
& 155 (73.11\%) & 211 (71.04\%) & 434 (73.06\%) & 149 (72.68\%) & 72
(72.73\%) & 212 (69.28\%) & 349 (81.92\%) \\
Indeterminate & 418 (40.31\%) & 113 (28.25\%) & 143 (33.26\%) & 120
(26.55\%) & 57 (26.89\%) & 86 (28.96\%) & 160 (26.94\%) & 56 (27.32\%) &
27 (27.27\%) & 94 (30.72\%) & 77 (18.08\%) \\
Total & 1,037 (100.00\%) & 400 (100.00\%) & 430 (100.00\%) & 452
(100.00\%) & 212 (100.00\%) & 297 (100.00\%) & 594 (100.00\%) & 205
(100.00\%) & 99 (100.00\%) & 306 (100.00\%) & 426 (100.00\%) \\

\end{longtable}

As previously mentioned, the macroscopic study was followed by a
petrographic approach. To better characterize the different chert types
identified in the macroscopic analysis and aid in minimizing the
inherent caveats of macroscopic methods, 22 thin sections were prepared
at the Thin Section Lab (Toul, France). The samples chosen for this
analysis include at least one sample from each type present throughout
the stratigraphy of the Terrace area, from different archaeological
levels (Table~\ref{tbl-TS-list}). For types that showed high macroscopic
variability (e.g.~translucency and colour variation or differentiated
bioclast facies), thin sections from sub-types were analysed to collect
petrographic data that could reflect this variability. Petrographic thin
sections were analysed by polarized light microscopy (Nikon LV100ND) to
determine the mineral composition, textural characteristics, allochems,
and bioclasts, as well as any other alterations.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0808}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1010}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0606}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.7576}}@{}}

\caption{\label{tbl-TS-list}List of archaeological thin sections, with
the summarised description of observed macroscopic variability and
sub-types.}

\tabularnewline
